---
title: "README"
output:
 github_document:
 pandoc_args: ["--wrap=none"]
always_allow_html: true
---
>Install the tidyverse library

```{r}
if (!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)

```
>We’ll start our exploration by visualizing the distribution of weights (carat) of ~54,000 diamonds from the diamonds dataset. Since carat is a numerical variable, we can use a histogram:

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.5)
```

>Let’s take a look at the distribution of carat for smaller diamonds.

```{r}
smaller <- diamonds |> 
  filter(carat < 3)

ggplot(smaller, aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

>Unusual Values
---
>Outliers are observations that are unusual; data points that don’t seem to fit the pattern. Sometimes outliers are data entry errors, sometimes they are simply values at the extremes that happened to be observed in this data collection, and other times they suggest important new discoveries. When you have a lot of data, outliers are sometimes difficult to see in a histogram. For example, take the distribution of the y variable from the diamonds dataset. The only evidence of outliers is the unusually wide limits on the x-axis.

```{r}
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
```


>There are so many observations in the common bins that the rare bins are very short, making it very difficult to see them (although maybe if you stare intently at 0 you’ll spot something). To make it easy to see the unusual values, we need to zoom to small values of the y-axis with coord_cartesian():

```{r}
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5) +
  coord_cartesian(ylim = c(0, 50))
```

>This allows us to see that there are three unusual values: 0, ~30, and ~60. We pluck them out with dplyr:

```{r}
unusual <- diamonds |> 
  filter(y < 3 | y > 20) |> 
  select(price, x, y, z) |>
  arrange(y)
unusual
#> # A tibble: 9 × 4
#>   price     x     y     z
#>   <int> <dbl> <dbl> <dbl>
#> 1  5139  0      0    0   
#> 2  6381  0      0    0   
#> 3 12800  0      0    0   
#> 4 15686  0      0    0   
#> 5 18034  0      0    0   
#> 6  2130  0      0    0   
#> 7  2130  0      0    0   
#> 8  2075  5.15  31.8  5.12
#> 9 12210  8.09  58.9  8.06
```
>EXCERCISES
---
>Exercise 10.3.3

>1.Explore the distribution of each of the x, y, and z variables in diamonds. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

`To explore distributions between x,y and z , we are going to plot histograms for each of the variables`
```{r}
# Load the necessary library
library(tidyverse)

# Plot histograms of x, y, and z
diamonds %>%
  select(x, y, z) %>%
  pivot_longer(everything(), names_to = "dimension", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~dimension, scales = "free") +
  labs(title = "Distributions of x, y, and z in diamonds dataset",
       x = "Value (mm)", y = "Count")
```
>Most values for x and y fall between 3 mm and 10 mm, suggesting they represent the length and width.
The z values are generally smaller, typically between 2 mm and 6 mm, which aligns with the depth of a diamond (top to bottom).
>Also notice that there are outliers in the data set with some diamonds having dimensions of 0 which is impractical
---

```{r}
# Boxplots to detect outliers visually
diamonds %>%
  select(x, y, z) %>%
  pivot_longer(cols = everything(), names_to = "dimension", values_to = "value") %>%
  ggplot(aes(x = dimension, y = value)) +
  geom_boxplot(outlier.color = "red", fill = "lightblue") +
  labs(title = "Boxplots of x, y, and z", y = "Value (mm)", x = "Dimension")
```

>2.	Explore the distribution of price. Do you discover anything unusual or surprising? (Hint: Carefully think about the binwidth and make sure you try a wide range of values.)

---
>Again we draw a histogram to visualize the distribution


```{r}
library(tidyverse)

# Histogram of diamond prices
ggplot(diamonds, aes(x = price)) +
  geom_histogram(bins =50, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Diamond Prices", x = "Price (USD)", y = "Count")
```

>The plot is highly right-skewed distribution. We can see that Most diamonds are priced under $5,000. A long tail extends toward high-priced diamonds, with some over $15,000.

>Lets increase binwidth and see

```{r}
# Histogram of diamond prices
ggplot(diamonds, aes(x = price)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Diamond Prices", x = "Price (USD)", y = "Count")
```



>3.How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

```{r}
library(tidyverse)

# Count diamonds with 0.99 and 1 carat
diamonds %>%
  filter(carat %in% c(0.99, 1)) %>%
  count(carat)
```
>There are far fewer diamonds listed as 0.99 carat than 1.00 carat.
This sharp drop suggests market behavior.Jewelers and cutters might cut slightly heavier stones down to exactly 1.00 carat to fetch higher prices.As a result, diamonds just under 1 carat (like 0.99) are undervalued and therefore less common in inventories.

>4.	Compare and contrast coord_cartesian() vs. xlim() or ylim() when zooming in on a histogram. What happens if you leave binwidth unset? What happens if you try and zoom so only half a bar shows?

```{r}
ggplot(diamonds, aes(x = price)) + 
  geom_histogram() + 
  xlim(0, 5000)+labs(title = "For data that is limited with xlim()")
```

>xlim()/ylim() filters data before plotting .Only the data within the limits is used to calculate bins.
If only part of a bar shows, the entire bar is dropped, because the data outside the limit is excluded
Binning is re-calculated — no partial bars allowed

```{r}
ggplot(diamonds, aes(x = price)) + 
  geom_histogram() + 
  coord_cartesian(xlim = c(0, 5000))

```

>EXCERCISE 10.4.1

>1. What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference in how missing values are handled in histograms and bar charts?

>Missing values are automatically dropped for a histogram.For a bar chart, if the factor variable itself has NA, the bar is plotted  unless you explicitly remove it

>2. What does na.rm = TRUE do in mean() and sum()?

>Removes NA values before computing the statistic.

>3. Recreate the frequency plot of scheduled_dep_time colored by whether the flight was cancelled or not. Also facet by the cancelled variable. Experiment with different values of the scales variable in the faceting function to mitigate the effect of more non-cancelled flights than cancelled flights.

>Load the dataset

```{r}
library(nycflights13)
```
>Now lets plot the frequency plot with scales set to  "free"

```{r}
# Add a 'cancelled' column: TRUE if dep_time is NA
flights <- flights %>%
  mutate(cancelled = is.na(dep_time))

# Plot histogram of scheduled departure time (24-hour format)
ggplot(flights, aes(x = sched_dep_time, fill = cancelled)) +
  geom_histogram(binwidth = 100, position = "identity", alpha = 0.6) +
  facet_wrap(~cancelled, scales = "free") +
  labs(
    title = "Scheduled Departure Times by Cancellation Status with scale set to free",
    x = "Scheduled Departure Time (hhmm)",
    y = "Count",
    fill = "Cancelled"
  ) +
  theme_minimal()
  
```

>Now lets plot the frequency plot with scales set to  "free_y"

```{r}
# Add a 'cancelled' column: TRUE if dep_time is NA
flights <- flights %>%
  mutate(cancelled = is.na(dep_time))

# Plot histogram of scheduled departure time (24-hour format)
ggplot(flights, aes(x = sched_dep_time, fill = cancelled)) +
  geom_histogram(binwidth = 100, position = "identity", alpha = 0.6) +
  facet_wrap(~cancelled, scales = "free_y") +
  labs(
    title = "Scheduled Departure Times by Cancellation Status with scale set to free_y",
    x = "Scheduled Departure Time (hhmm)",
    y = "Count",
    fill = "Cancelled"
  ) +
  theme_minimal()
  
```

>Now lets plot the frequency plot with scales set to  "free_x"

```{r}
# Add a 'cancelled' column: TRUE if dep_time is NA
flights <- flights %>%
  mutate(cancelled = is.na(dep_time))

# Plot histogram of scheduled departure time (24-hour format)
ggplot(flights, aes(x = sched_dep_time, fill = cancelled)) +
  geom_histogram(binwidth = 100, position = "identity", alpha = 0.6) +
  facet_wrap(~cancelled, scales = "free_x") +
  labs(
    title = "Scheduled Departure Times by Cancellation Status with scale set to free_x",
    x = "Scheduled Departure Time (hhmm)",
    y = "Count",
    fill = "Cancelled"
  ) +
  theme_minimal()
  
```
>Now lets plot the frequency plot with scales set to  "fixed"

```{r}
# Add a 'cancelled' column: TRUE if dep_time is NA
flights <- flights %>%
  mutate(cancelled = is.na(dep_time))

# Plot histogram of scheduled departure time (24-hour format)
ggplot(flights, aes(x = sched_dep_time, fill = cancelled)) +
  geom_histogram(binwidth = 100, position = "identity", alpha = 0.6) +
  facet_wrap(~cancelled, scales = "free_x") +
  labs(
    title = "Scheduled Departure Times by Cancellation Status with scale set to fixed",
    x = "Scheduled Departure Time (hhmm)",
    y = "Count",
    fill = "Cancelled"
  ) +
  theme_minimal()
  
```

  